Trends from 30-Year Observations of Downward Solar Irradiance in Thessaloniki, Greece

### 2024-01-01 MDPI_submission.Rmd
Commit 2024-01-01 19:12

# institutionalreview: |
#   In this section, you should add the Institutional Review Board Statement and
#   approval number, if relevant to your study. You might choose to exclude
#   this statement if the study did not require ethical approval. Please note
#   that the Editorial Office might ask you for further information. Please add
#   “The study was conducted in accordance with the Declaration of Helsinki,
#   and approved by the Institutional Review Board (or Ethics Committee) of
#   NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies
#   involving humans. OR “The animal study protocol was approved by the
#   Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE
#   (protocol code XXX and date of approval).” for studies involving animals.
#   OR “Ethical review and approval were waived for this study due to REASON
#   (please provide a detailed justification).” OR “Not applicable” for
#    studies not involving humans or animals.

# informedconsent: |
#   Any research article describing a study involving humans should contain this
#   statement. Please add ``Informed consent was obtained from all subjects
#   involved in the study.'' OR ``Patient consent was waived due to REASON
#   (please provide a detailed justification).'' OR ``Not applicable'' for
#   studies not involving humans. You might also choose to exclude this statement
#   if the study did not involve humans.
#
#   Written informed consent for publication must be obtained from participating
#   patients who can be identified (including by the patients themselves). Please
#   state ``Written informed consent has been obtained from the patient(s) to
#   publish this paper'' if applicable.

@@

# conflictsofinterest: |
#   Declare conflicts of interest or state 'The authors declare no conflict of
#   interest.' Authors must identify and declare any personal circumstances or
#   interest that may be perceived as inappropriately influencing the
#   representation or interpretation of reported research results. Any role of the
#   funding sponsors in the design of the study; in the collection, analyses or
#   interpretation of data in the writing of the manuscript, or in the decision to
#   publish the results must be declared in this section. If there is no role,
#   please state 'The founding sponsors had no role in the design of the study;
#   in the collection, analyses, or interpretation of data; in the writing of the
#   manuscript, an in the decision to publish the results'.

@@

```{r echo=F, include=T, warning=FALSE, message=FALSE}
suppressMessages(
    res <- rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "70%"   )
knitr::opts_chunk$set(fig.align  = "center")
knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
# source("./DHI_GHI_00_raw_data.R")
# load(common_data)
load("./data/Input_1_longterm_trends.Rda")
## number of measurements for each dataset
SDR_N_All   <- sum(  ALL_1_daily_DESEAS[,GLB_att_N], na.rm = T)
SDR_N_Clear <- sum(CLEAR_1_daily_DESEAS[,GLB_att_N], na.rm = T)
SDR_N_Cloud <- sum(CLOUD_1_daily_DESEAS[,GLB_att_N], na.rm = T)

@@

    # else {
    #     
    #     output <- vector(mode = "character", length = options$fig.num + 1)
    #     
    #     for (i in 1:options$fig.num) {
    #         output[i] <- sprintf("\\includegraphics{%s}", fig_path(number = i))
    #     }
    #     
    #     output[i+1] <- "\\end{adjustwidth}"
    #     return(paste(output, collapse = ""))
    # }
})

@@

<!--
Aerosols interactions

#Yang2021  Cloud “shrinking” and “optical thinning” in the “dimming” period and a subsequent recovery in the “brightening” period over China

Aerosols not only strengthen but also weaken the
growth of clouds with respect to their coverage and
optical thickness, depending on the levels of pollution
and the associated amounts of aerosols as postulated
in a conceptual framework (Wild 2009a, 2012, Yang
et al 2012).

#Wild2012 : simplify interaction nice graph
-->

@@

# TSI$cor.null_value     <- NULL
# TSI$cor.alternative    <- NULL
# TSI$cor.method         <- NULL
# TSI$cor.data_name      <- NULL
# TSI$cor.p              <- NULL
TSI$cor.conf_level      <- NULL
care <- grep("^cor\\.", names(TSI), value = T)
for (an in care) {
    TSI[, (an) := NULL ]
}

@@

In order to unveil further the features of the variability of the three
datasets, Figure\ \ref{fig:cusumnotrendmonthly}
presents another set of CUSUM plots using anomalies
after the long-term linear trend is removed. With this approach, periods
when the CUSUMs diverge from zero can be interpreted as a systematic
variation of SDR from the climatological mean. When the CUSUM is
increasing, the anomalies values are above the climatology and vice
versa. 
Overall, for all- and cloudy-sky conditions (Figures\ \ref{fig:cusumnotrendmonthly-1} and\ \ref{fig:cusumnotrendmonthly-3}) we
observe periods with anomalies diverging from the climatological values,
each lasting for several years. These fluctuations are probably within
the natural variability, and no distinct changes are identified. The
pattern in both datasets is similar, suggesting prevalence in cloudy
skies over Thessaloniki.
For clear skies (Figure\ \ref{fig:cusumnotrendmonthly-2}) the distinct change
in 2004 is now clearer. The most likely reason for this change is the
monotonic reduction of aerosols in Thessaloniki. In that year, there was a
change in the rate of decrease in aerosol optical depth, as illustrated
in Figure 7 of @Siomos2020.
This abrupt change in CUSUMs lasted
until about 2010, when the anomalies become again variable.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            % \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            % \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
        \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{adjustwidth}
\end{figure}

@@

The solar zenith angle is a major factor affecting the SDR, since
increases in SZA leads to enhancement of the radiation path in the
atmosphere, especially in urban environments with human activities
emitting aerosols 
[@Wang2021].
In order to estimate the effect of SZA on the SDR trends, we grouped the data in bins of $1^\circ$ SZA, and calculated the overall trend for each bin, separately for the daily
periods before noon and after noon (Figure\ \ref{fig:szatrends}).
Although there are seasonal dependencies of the minimum SZA (see Appendix, Figure\ \ref{fig:SZAtrendSeason}), these dependencies are not discussed further.

For all-sky conditions the brightening effect of SDR (positive trend) increases with SZAs (Figures\ \ref{fig:szatrends-1})
ranging from about $0.1\,\%/y$ to
about $0.7\,\%/y$ for the statistically significant trends.
The trends in the morning and afternoon hours are more or
less consistent with small differences at small SZAs which can be
attributed to effects on clear sky SDR from systematic diurnal patterns
of aerosols during the warm period of the year, consistently with the results reported for China by @Wang2021. 
Note that SZAs less than $25^\circ$ can only occur during the warm period of the year around noon when clear skies are more
frequent. 
The increasing trend with SZA is likely caused by the
increased attenuation of SDR with SZA. The effect is larger when aerosol
and/or cloud layers are optically thicker, therefore, decreases in
aerosol and clouds through the study period will result in larger
positive trends of SDR at larger SZAs.

Under clear skies (Figure\ \ref{fig:szatrends-2}), the trends are smaller and less variable,
ranging between $0.1$ and $0.15\,\%/y$ up to $77^\circ$ SZA.
At higher SZAs and in
the afternoon hours, there is a sharp increase in the trend up to $0.3\,\%/y$,
which may have been caused by the long path length of radiation through
the atmosphere as discussed above for the all-sky conditions. The small
differences in the trend between morning and afternoon between $35^\circ$ and
$60^\circ$ SZA is likely a result of less attenuation of SDR in the morning
hours due to lesser amounts of aerosols and a shallower boundary layer.

For cloudy-sky conditions (Figure\ \ref{fig:szatrends-3}), we cannot discern any significant dependence of the SDR trend with SZA as the variability of irradiance is
dominated by the cloud effects leading to insignificant
trends. Statistically significant trends appear only in the afternoon and
for SZAs larger than $60^\circ$.
The sharp increase of the trend at SZAs
larger than $\sim{75}^{\circ}$, observed also for clear skies, is
probably associated with stronger attenuation by clouds under oblique
incidence angles, which result also in smaller variability.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            \subfloat[All skies.\label{fig:szatrends-1}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-1}}\hfill
            \subfloat[Clear skies.\label{fig:szatrends-2}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-4}}\hfill
            \subfloat[Cloudy skies.\label{fig:szatrends-3}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-7}}
        }
        \caption{Long term trends of daily SDR as a function of SZA for (a) all-sky, (b) clear-sky and (c) cloudy-sky conditions, separately for morning and afternoon periods. Solid shapes represent statistically significant trends ($p < 0.005$).}\label{fig:szatrends}
    \end{adjustwidth}
\end{figure}

<!-- ```{r szatrends, echo = FALSE, fig.cap='Long term trends of SDR as a function of SZA separately from morning and afternoon periods. Solid shapes  represent statistically significant trends ($p < 0.005$).', fig.subcap=c('All skies.', 'Clear skies.', 'Cloudy skies.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics("./images/SzaTrends-1.png") -->
<!-- knitr::include_graphics("./images/SzaTrends-4.png") -->
<!-- knitr::include_graphics("./images/SzaTrends-7.png") -->
<!-- ``` -->

<!-- ```{r szatrendsmonthly, echo = FALSE, fig.cap='*** Montlhly *** Distribution of the SDR\'s long term trends by SZA before and after noon. The solid data points represent values with an acceptable statistical significance ($p<0.005$).', fig.subcap=c('All-sky conditions.', 'Clear-sky conditions.', 'Cloud-sky conditions.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3} -->
<!-- knitr::include_graphics("./images/SzaTrendsMonthly-1.png") -->
<!-- knitr::include_graphics("./images/SzaTrendsMonthly-4.png" ) -->
<!-- knitr::include_graphics("./images/SzaTrendsMonthly-7.png") -->
<!-- ``` -->

## Long term SDR trends by season

@@

## numeric precision
spp$Tlm_Std..Error  <- round(spp$Tlm_Std..Error * 2, 2)
spp$slope           <- round(spp$slope,              2)
spp$slope.p         <- round(spp$slope.p,            3)
spp$cor.estimate    <- round(spp$cor.estimate,       2)
# spp$slope.stat_sig  <- signif(spp$slope.stat_sig, 3)
# spp$Rsqrd           <- signif(spp$Rsqrd,          3)

# spp$Tlm_Pr...t..

@@

Similarly to the long term trends from daily means of SDR discussed above, we have calculated the trend for the three sky conditions and for each season of the year, using the corresponding mean monthly anomalies
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
Table\ \ref{tab:trendseasontable} contains also the $2\sigma$ standard error, the Pearson's correlation coefficient R and the corresponding p-value.
The winter linear trends exhibit generally the largest R values ranging between $0.54$ and $0.60\,\%/y$. 
<!-- all sky -->
For all-sky conditions the trend in SDR in winter is the largest
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$),
followed by the trend in autumn
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$,
a value close to the long term trend) both statistically significant at the $95\,\%$ confidence level.
In spring and summer, the trends are much smaller and of lesser statistical significance.
These seasonal differences indicate a possible relation of the trends in SDR to trends in clouds during winter and autumn.
<!-- clear sky -->
For clear-skies, the trend in winter is $`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$ and is 
associated with the decreasing trend in
aerosol optical depth [@Siomos2020]. Moreover, it is almost half of that for
all-skies, which is another indication of a decreasing trend in cloud
optical thickness. 
<!-- clear sky -->
In other seasons the clear-sky trend is very small
(below $0.1\,\%/y$). 
<!-- cloud dsky -->
Finally, for cloudy-skies the
winter trend is the largest 
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$)
and greater than for all-skies,
followed by a much smaller trend in autumn 
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$).
<!-- ($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$) -->

@@

<!-- % \includegraphics[width=1\linewidth]{./images/SeasonalTrendsTogether3-2}  % Seasonal from Daily -->

<!-- ```{r seasonalALL, echo=F, fig.cap="Linear trends (black lines) of monthly mean anomalies of SDR by season (rows of plots) for the three sky conditions (columns of plots).", out.width="100%"} -->
<!-- knitr::include_graphics("./images/SeasonalTrendsTogether3-2.png") -->
<!-- ``` -->

<!-- ```{r seasonalALL, echo=F, fig.cap="Trends by season for all-sky condition. Displays monthly means of daily means."} -->
<!-- knitr::include_graphics("./images/SeasonalTrends-1.png") -->
<!-- ``` -->

<!-- ```{r seasonalCLEAR, echo=F, fig.cap="Trends by season for clear-sky condition. Displays monthly means of daily means."} -->
<!-- knitr::include_graphics("./images/SeasonalTrends-2.png") -->
<!-- ``` -->

<!-- ```{r seasonalCLOUD, echo=F, fig.cap="Trends by season for Cloudy-sky condition. Displays monthly means of daily means."} -->
<!-- knitr::include_graphics("./images/SeasonalTrends-3.png") -->
<!-- ``` -->

```{r trendseasontable, echo=F}

spp2 <- spp

@@

# sub("Trend S.E.", "Trend S.E. ($2\\sigma$)", names(spp2))

# panderOptions("table.alignment.default", "left")
# panderOptions("table.alignment.default", "center")
# pander(spp,
#        justify = c('center', 'right', 'right', 'center'),
#        split.table = Inf,
#        caption = "\\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.")

## use kable
x_latex <- knitr::kable(spp2,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = FALSE,
                        col.names   = c("Sky condition",
                                        "Season",
                                        "Trend [\\%/year]",
                                        "Trend S.E. ($2\\sigma$)",
                                        "Pearson correl.",
                                        "Trend p-value" ),
                        # format.args = list(digits = 3),
                        linesep     = "",
                        align       = "ccrrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        # position    = "H",
                        caption  = "SDR linear trends of monthly anomalies for each season of the year and related statistical parameters.")

# x_latex %>%
#     collapse_rows(columns = 1, valign = "middle", latex_hline = "custom") %>%
#     kable_styling(latex_options = c("striped",
#                                     "hold_position")
#     )

>>>>

institutionalreview: |
  Not applicable.

informedconsent: |
  Not applicable.

@@

conflictsofinterest: |
  The authors declare no conflict of interest.

@@

```{r echo=F, include=T, warning=FALSE, message=FALSE}
suppressMessages(
    res <- rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "70%"   )
knitr::opts_chunk$set(fig.align  = "center")
knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
load("./data/Input_1_longterm_trends.Rda")
## number of measurements for each dataset
SDR_N_All   <- sum(  ALL_1_daily_DESEAS[,GLB_att_N], na.rm = T)
SDR_N_Clear <- sum(CLEAR_1_daily_DESEAS[,GLB_att_N], na.rm = T)
SDR_N_Cloud <- sum(CLOUD_1_daily_DESEAS[,GLB_att_N], na.rm = T)

@@

    # else {
    #     output <- vector(mode = "character", length = options$fig.num + 1)
    #     
    #     for (i in 1:options$fig.num) {
    #         output[i] <- sprintf("\\includegraphics{%s}", fig_path(number = i))
    #     }
    #     
    #     output[i+1] <- "\\end{adjustwidth}"
    #     return(paste(output, collapse = ""))
    # }
})

@@

@@

TSI$cor.conf_level      <- NULL
care <- grep("^cor\\.", names(TSI), value = T)
for (an in care) {
    TSI[, (an) := NULL ]
}

@@

In order to unveil further the features of the variability of the three
datasets, Figure\ \ref{fig:cusumnotrendmonthly}
presents another set of CUSUM plots using anomalies
after the long-term linear trend is removed. With this approach, periods
when the CUSUMs diverge from zero can be interpreted as a systematic
variation of SDR from the climatological mean. When the CUSUM is
increasing, the anomalies values are above the climatology, and vice
versa. 
Overall, for all- and cloudy-sky conditions (Figure\ \ref{fig:cusumnotrendmonthly}a,c), we
observe periods with anomalies diverging from the climatological values,
each lasting for several years. These fluctuations are probably within
the natural variability, and no distinct changes are identified. The
pattern in both datasets is similar, suggesting prevalence in cloudy
skies over Thessaloniki.
For clear skies (Figure\ \ref{fig:cusumnotrendmonthly}b), the distinct change
in 2004 is now clearer. The most likely reason for this change is the
monotonic reduction of aerosols in Thessaloniki. In that year, there was a
change in the rate of decrease in aerosol optical depth, as illustrated
in Figure 7 of @Siomos2020.
This abrupt change in CUSUMs lasted
until about 2010, when the anomalies become again variable.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            % \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            % \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
    \end{adjustwidth}
    \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

The solar zenith angle is a major factor affecting the SDR, as
increases in SZA leads to enhancement of the radiation path in the
atmosphere, especially in urban environments with human activities
emitting aerosols [@Wang2021].
In order to estimate the effect of SZA on the SDR trends, we grouped the data in 
bins of $1^\circ$ SZA and calculated the overall trend for each bin separately for the daily
periods before noon and after noon (Figure\ \ref{fig:szatrends}).
Although there are seasonal dependencies of the minimum SZA (see Appendix \ref{appendix}, Figure\ \ref{fig:SZAtrendSeason}), these dependencies are not discussed further.

For all-sky conditions, the brightening effect of SDR (positive trend) increases with SZAs (Figure\ \ref{fig:szatrends}a),
ranging from about $0.1\,\%/y$ to
about $0.7\,\%/y$ for the statistically significant trends.
The trends in the morning and afternoon hours are more or
less consistent with small differences at small SZAs, which can be
attributed to effects on clear sky SDR from systematic diurnal patterns
of aerosols during the warm period of the year, consistent with the 
results reported for China by @Wang2021. 
Note that SZAs less than $25^\circ$ can only occur during the warm period
of the year around noon when clear skies are more frequent. 
The increasing trend with SZA is likely caused by the
increased attenuation of SDR with SZA. The effect is larger when aerosol
and/or cloud layers are optically thicker; therefore, decreases in
aerosol and clouds through the study period will result in larger
positive trends of SDR at larger SZAs.

Under clear skies (Figure\ \ref{fig:szatrends}b), the trends are smaller
and less variable,
ranging between $0.1$ and $0.15\,\%/y$ up to $77^\circ$ SZA.
At higher SZAs and in
the afternoon hours, there is a sharp increase in the trend up to $0.3\,\%/y$,
which may have been caused by the long path length of radiation through
the atmosphere as discussed above for the all-sky conditions. The small
differences in the trend between morning and afternoon between $35^\circ$ and
$60^\circ$ SZA is likely a result of less attenuation of SDR in the morning
hours due to lesser amounts of aerosols and a shallower boundary layer.

For cloudy-sky conditions (Figure\ \ref{fig:szatrends}c), we cannot discern any significant dependence of the SDR trend with SZA, as the variability of irradiance is
dominated by the cloud effects leading to insignificant
trends. Statistically significant trends appear only in the afternoon and
for SZAs larger than $60^\circ$.
The sharp increase of the trend at SZAs
larger than $\sim{75}^{\circ}$, observed also for clear skies, is
probably associated with stronger attenuation by clouds under oblique
incidence angles, which also result in smaller variability.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:szatrends-1}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-1}}\hfill
            % \subfloat[Clear skies.\label{fig:szatrends-2}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-4}}\hfill
            % \subfloat[Cloudy skies.\label{fig:szatrends-3}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-7}}
        }
    \end{adjustwidth}
    \caption{Long-term trends of daily SDR as a function of SZA for (a) all-sky, (b) clear-sky and (c) cloudy-sky conditions, separately for morning and afternoon periods. Solid shapes represent statistically significant trends ($p < 0.005$).}\label{fig:szatrends}
\end{figure}

## Long-Term SDR Trends by Season

@@

## numeric precision
spp$Tlm_Std..Error  <- round(spp$Tlm_Std..Error * 2, 2)
spp$slope           <- round(spp$slope,              2)
spp$slope.p         <- round(spp$slope.p,            3)
spp$cor.estimate    <- round(spp$cor.estimate,       2)

@@

Similarly to the long term trends from daily means of SDR discussed above,
we have calculated the trend for the three sky conditions and for each 
season of the year, using the corresponding mean monthly anomalies
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
Table\ \ref{tab:trendseasontable} also contains the $2\sigma$ standard error,
the Pearson's correlation coefficient R and the corresponding p-value.
The winter linear trends exhibit generally the largest R values ranging
between $0.54$ and $0.60\,\%/y$. 
<!-- all sky -->
For all-sky conditions, the trend in SDR in winter is the largest
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$),
followed by the trend in autumn
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$,
a value close to the long term trend) both statistically significant at the $95\,\%$ confidence level.
In spring and summer, the trends are much smaller and of lesser statistical significance.
These seasonal differences indicate a possible relation of the trends in SDR to trends in clouds during winter and autumn.
<!-- clear sky -->
For clear-skies, the trend in winter is $`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$ and is 
associated with the decreasing trend in
aerosol optical depth [@Siomos2020]. Moreover, it is almost half of that for
all-skies, which is another indication of a decreasing trend in cloud
optical thickness. 
<!-- clear sky -->
In other seasons the clear-sky trend is very small (below $0.1\,\%/y$). 
<!-- cloud dsky -->
Finally, for cloudy-skies the winter trend is the largest 
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$)
and greater than for all-skies, followed by a much smaller trend in autumn 
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$).

@@

```{r trendseasontable, echo=F}
spp2 <- spp

@@

x_latex <- knitr::kable(spp2,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = FALSE,
                        col.names   = c("Sky condition",
                                        "Season",
                                        "Trend [\\%/year]",
                                        "Trend S.E. ($2\\sigma$)",
                                        "Pearson correl.",
                                        "Trend p-value" ),
                        # format.args = list(digits = 3),
                        linesep     = "",
                        align       = "ccrrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        # position    = "H",
                        caption  = "SDR linear trends of monthly anomalies for each season of the year and related statistical parameters.")
